---
title: "*collapse* and *sf*"
subtitle: "Fast Manipulation of Simple Features Data Frames"
author: "Sebastian Krantz, with a contribution by Grant McDermott"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true

vignette: >
  %\VignetteIndexEntry{collapse and sf: Fast Manipulation of Simple Features Data Frames}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
pre {
  max-height: 900px;
  overflow-y: auto;
}

pre[class] {
  max-height: 900px;
}
```

```{r, echo = FALSE, message = FALSE, warning=FALSE}

NMAC <- Sys.getenv("NMAC") == "TRUE"
if(NMAC) library(sf)
library(microbenchmark)
library(collapse)
knitr::opts_chunk$set(error = FALSE, message = FALSE, warning = FALSE, 
                      comment = "#", tidy = FALSE, cache = FALSE, collapse = TRUE,
                      fig.width = 8, fig.height = 5, 
                      out.width = '100%')

oldopts <- options(width = 100L)
set.seed(101)
```
<!--
*collapse* is a C/C++ based package for data transformation and statistical computing in R. It's aims are:

1. To facilitate complex data transformation, exploration and computing tasks in R.
2. To help make R code fast, flexible, parsimonious and programmer friendly. 
-->
This short vignette focuses on using *collapse* with the popular *sf* package by Edzer Pebesma. It shows that *collapse* supports easy manipulation of *sf* data frames, at computation speeds far above *dplyr*. 

*collapse* is class-agnostic and can theoretically be used to manipulate any kind of list, matrix or data frame based object. Previous versions of *collapse* could manipulate *sf* data frames, but required users to manually include the geometry column when using functions like in `fselect` and `fsubset`, and exclude it in other functions like `qsu`. 

Version 1.6.0 of the package adds internal support for *sf* data frames where the geometry column is handled by most essential functions (`fselect`, `get_vars`, `num_vars`, `fsubset`, `ss`, `fgroup_by`, `qsu`, `descr`, `varying`, `funique`, `roworder`, `rsplit`, `fcompute`, ...). This largely does not affect previously written code where the geometry column is explicitly selected. To demonstrate these features, we can load a test dataset provided by *sf*:

```{r, eval=NMAC}
library(collapse)
library(sf)

nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
options(sf_max_print = 3)
nc
```

## Summarising sf Data Frames

The first thing that *collapse* 1.6.0 adds is summary statistics for *sf* data frames that automatically exclude the 'geometry' column:

```{r, eval=NMAC}
# Which columns have at least 2 non-missing distinct values
varying(nc) 

# Quick summary stats
qsu(nc)

# Detailed statistics description of each column
descr(nc)
```


## Selecting Columns and Subsetting

We can now also easily select columns from the *sf* data frame without having to worry about taking along 'geometry':

```{r, eval=NMAC}
# Selecting a sequence of columns
fselect(nc, AREA, NAME:FIPSNO)

# Same using standard evaluation (gv is a shorthand for get_vars())
gv(nc, c("AREA", "NAME", "FIPS", "FIPSNO"))
```

The same applies to subsetting rows (and columns):

```{r, eval=NMAC}
# A fast and enhanced version of base::subset, now also supporting sf data 
fsubset(nc, AREA > fmean(AREA), AREA, NAME:FIPSNO)

# A fast version of `[` (where i is used and optionally j)
ss(nc, 1:10, c("AREA", "NAME", "FIPS", "FIPSNO"))
```

This is significantly faster than using `[`, `dplyr::select` or `dplyr::filter`: 

```{r, eval=NMAC}
library(microbenchmark)
library(dplyr)

# Selecting columns
microbenchmark(collapse = fselect(nc, AREA, NAME:FIPSNO), 
               dplyr = select(nc, AREA, NAME:FIPSNO),
               collapse2 = gv(nc, c("AREA", "NAME", "FIPS", "FIPSNO")), 
               sf = nc[c("AREA", "NAME", "FIPS", "FIPSNO")])
# Subsetting
microbenchmark(collapse = fsubset(nc, AREA > fmean(AREA), AREA, NAME:FIPSNO), 
               dplyr = select(nc, AREA, NAME:FIPSNO) %>% filter(AREA > fmean(AREA)),
               collapse2 = ss(nc, 1:10, c("AREA", "NAME", "FIPS", "FIPSNO")), 
               sf = nc[1:10, c("AREA", "NAME", "FIPS", "FIPSNO")])

```

It needs to be noted that at this point *collapse* functions don't subset the 'agr' attribute on selecting columns, which (if specified) relates columns (attributes) to the geometry, and also don't modify the 'bbox' attribute giving the overall boundaries of a set of geometries when subsetting the *sf* data frame. Keeping the full 'agr' attribute is not problematic for all practical purposes, but not changing 'bbox' upon subsetting may lead to too large margins when plotting the geometries of a subsetted *sf* data frame. 

## Aggregation and Grouping

The flexibility and speed of `collap` for aggregation can be used on *sf* data frames. A separate method for *sf* objects was not considered necessary as one can simply aggregate the geometry column using `st_union`:

```{r, eval=NMAC}
# Aggregating by variable SID74 using the median for numeric and the mode for categorical columns
collap(nc, ~ SID74, custom = list(fmedian = is.numeric, 
                                  fmode = is.character, 
                                  st_union = "geometry")) # or use is.list to fetch the geometry
```

*sf* data frames can also be grouped and then aggregated using `fsummarise`:

```{r, eval=NMAC}
nc %>% fgroup_by(SID74)

nc %>% fgroup_by(SID74) %>%
  fsummarise(AREA_Ag = fsum(AREA), 
             Perimeter_Ag = fmedian(PERIMETER),
             geometry = st_union(geometry))
```

It needs to be noted here that typically most of the time in aggregation is consumed by `st_union` so that the speed of *collapse* does not really become visible on most datasets. One exception is spatial panel data when aggregating over the time-dimension with identical geometries for each id over time. Such panels can quickly be aggregated using `ffirst` or `flast` to aggregate the geometry:

```{r, eval=NMAC}
# Creating a panel-dataset by simply duplicating nc for 2 different years
pnc <- list(`2000` = nc, `2001` = nc) %>% unlist2d("Year") %>% copyMostAttrib(nc)
pnc 

# Aggregating by NAME, using the last value for all categorical data
collap(pnc, ~ NAME, fmedian, catFUN = flast, cols = -1L)

# Using fsummarise to aggregate just two variables and the geometry
pnc_ag <- pnc %>% fgroup_by(NAME) %>%
  fsummarise(AREA_Ag = fsum(AREA), 
             Perimeter_Ag = fmedian(PERIMETER),
             geometry = flast(geometry))

# The geometry is still valid...
plot(slt(pnc_ag, AREA_Ag))

```


## Unique Values, Ordering, Splitting, Binding

Functions `funique` and `roworder(v)` work on sf data frames and ignore the geometry column for determining the unique values / order of rows. `rsplit` can be used to (recursively) split an *sf* data frame into multiple chunks. 

```{r, eval=NMAC}
# Splitting by SID74
rsplit(nc, ~ SID74) %>% head(2)
```

It should be noted here that the default in `rsplit` for data frames is `simplify = TRUE`, which for a single LHS variable would just split the column-vector. This does not apply to *sf* data frames as the 'geometry' column is always selected as well.

```{r, eval=NMAC}
# Only splitting Area
rsplit(nc, AREA ~ SID74) %>% head(1)

# For data frames the default simplify = TRUE drops the data frame structure
rsplit(qDF(nc), AREA ~ SID74) %>% head(1)
```

*sf* data frames can be combined using `rowbind()`, which will preserve the attributes of the first object. 

```{r, eval=NMAC}
# Splitting by each row and recombining
nc_combined <- nc %>% rsplit(seq_row(.)) %>% rowbind() 
identical(nc, nc_combined)
```

## Transformations

For transforming and computing columns, `ftransform(v)` and `settransform(v)` apply as to any other data frame. 

```{r, eval=NMAC}
ftransform(nc, scaled_AREA = fscale(AREA), 
               gsum_AREA = fsum(AREA, SID74, TRA = "replace_fill"))

# settransform materializes the change, same as nc <- ftransform(nc, ...)
settransform(nc, scaled_AREA = fscale(AREA), 
                 gsum_AREA = fsum(AREA, SID74, TRA = "replace_fill"))
nc
```

Special attention to *sf* data frames is afforded by `fcompute`, which can be used to compute new columns dropping existing ones - except for the geometry column and any columns selected through the `keep` argument. 

```{r, eval=NMAC}
fcompute(nc, scaled_AREA = fscale(AREA), 
             gsum_AREA = fsum(AREA, SID74, TRA = "replace_fill"), 
         keep = .c(AREA, SID74))
```

## Conversion to and from *sf*

The quick converters `qDF`, `qDT` and `qTBL` can be used to very efficiently convert *sf* data frames to standard data frames, *data.table*'s or *tibbles*, and the result can be converted back to the original *sf* data frame using `setAttrib`, `copyAttrib` or `copyMostAttrib`. 

```{r, eval=NMAC}
library(data.table)
qDT(nc)[, list(roll_AREA = frollmean(AREA, 2), geometry), by = SID74] %>% copyMostAttrib(nc)
```



## Conclusion

*collapse* provides no deep integration with the *sf* ecosystem, but sufficient features and flexibility to painlessly manipulate *sf* data frames with much greater performance. 


```{r, echo=FALSE}
options(oldopts)
```


