---
title: "*collapse* and *dplyr*"
subtitle: "Fast (Weighted) Aggregations, Transformations and Panel Computations in a Piped Workflow"
author: "Sebastian Krantz"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true

vignette: >
  %\VignetteIndexEntry{collapse and dplyr: Fast (Weighted) Aggregations, Transformations and Panel Computations in a Piped Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  
    <!--  %\VignetteIndexEntry{collapse and dplyr: Fast (weighted) Aggregations, Transformations and Time-Series/Panel Computations in a dplyr Workflow} -->
---

```{r, echo = FALSE, message = FALSE, warning=FALSE}
library(dplyr)
library(microbenchmark)
library(collapse)
knitr::opts_chunk$set(error = FALSE, message = FALSE, warning = FALSE, 
                      comment = "#", tidy = FALSE, cache = FALSE, collapse = TRUE,
                      fig.width = 8, fig.height = 5, 
                      out.width = '100%')

# knitr::opts_chunk$set(
#   comment = "#",
#     error = FALSE,
#      tidy = FALSE,
#     cache = FALSE,
#  collapse = TRUE,
#  fig.width = 8, 
#  fig.height= 5,
#  out.width='100%'
# )

NCRAN <- identical(Sys.getenv("NCRAN"), "TRUE")

oldopts <- options(width = 100L)
set.seed(101)
```

*collapse* is a C/C++ based package for data manipulation in R. It's aims are 

1. to facilitate complex data transformation and exploration tasks and 

2. to help make R code fast, flexible, parsimonious and programmer friendly.

This vignette focuses on the integration of *collapse* and the popular *dplyr* package by Hadley Wickham. In particular it will demonstrate how using *collapse*'s fast functions and some fast alternatives for *dplyr* verbs can substantially facilitate and speed up basic data manipulation, grouped and weighted aggregations and transformations, and panel-data computations (i.e. between- and within-transformations, panel-lags, differences and growth rates) in a *dplyr* (piped) workflow. 

***

**Notes:**

- This vignette is targeted at *dplyr* / *tidyverse* users. *collapse* is a standalone package and can be programmed efficiently without pipes or *dplyr* verbs. 

- The 'Introduction to *collapse*' vignette provides a thorough introduction to the package and a built-in structured documentation is available under `help("collapse-documentation")` after installing the package. In addition `help("collapse-package")` provides a compact set of examples for quick-start. 

***

## 1. Fast Aggregations

A key feature of *collapse* is it's broad set of *Fast Statistical Functions* (`fsum, fprod, fmean, fmedian, fmode, fvar, fsd, fmin, fmax, ffirst, flast, fNobs, fNdistinct`) which are able to substantially speed-up column-wise, grouped and weighted computations on vectors, matrices or data.frame's. The functions are S3 generic, with a default (vector), matrix and data.frame method, as well as a grouped_df method for grouped tibbles used by dplyr. The grouped tibble method has the following arguments:  

```{r eval=FALSE}
FUN.grouped_df(x, [w = NULL,] TRA = NULL, [na.rm = TRUE,]
               use.g.names = FALSE, keep.group_vars = TRUE, [keep.w = TRUE,] ...)
```

where `w` is a weight variable (available only to `fsum, fprod, fmean, fmode, fvar` and `fsd`), and `TRA` and can be used to transform `x` using the computed statistics and one of 10 available transformations (`"replace_fill", "replace", "-", "-+", "/", "%", "+", "*", "%%", "-%%"`). These transformations perform grouped replacing or sweeping out of the statistics computed by the function (discussed in section 2). `na.rm` efficiently removes missing values and is `TRUE` by default. `use.g.names` generates new row-names from the unique combinations of groups (default: disabled), whereas `keep.group_vars` (default: enabled) will keep the grouping columns as is custom in the native `data %>% group_by(...) %>% summarize(...)` workflow in *dplyr*. Finally, `keep.w` regulates whether a weighting variable used is also aggregated and saved in a column. For `fsum, fmean, fvar and fsd` this will compute the sum of the weights in each group, whereas `fmode` will return the maximum weight (corresponding to the mode) in each group and `fprod` returns the product of the weights. 

With that in mind, let's consider some straightforward applications.

### 1.1 Simple Aggregations

Consider the Groningen Growth and Development Center 10-Sector Database included in *collapse* and introduced in the main vignette:

```{r}
library(collapse)
head(GGDC10S)

# Summarize the Data: 
# descr(GGDC10S, cols = is.categorical)
# aperm(qsu(GGDC10S, ~Variable, cols = is.numeric))
```

Simple column-wise computations using the fast functions and pipe operators are performed as follows:

```{r}
library(dplyr)

GGDC10S %>% fNobs                       # Number of Observations
GGDC10S %>% fNdistinct                  # Number of distinct values
GGDC10S %>% select_at(6:16) %>% fmedian # Median
GGDC10S %>% select_at(6:16) %>% fmean   # Mean
GGDC10S %>% fmode                       # Mode
GGDC10S %>% fmode(drop = FALSE)         # Keep data structure intact
```

Moving on to grouped statistics, we can compute the average value added and employment by sector and country using:

```{r}
GGDC10S %>% 
  group_by(Variable,Country) %>%
  select_at(6:16) %>% fmean

```

Similarly we can aggregate using any other of the above functions.

<!-- ```{r} -->
<!-- GGDC10S %>%  -->
<!--   group_by(Variable,Country) %>% -->
<!--   select_at(6:16) %>% fmedian -->

<!-- GGDC10S %>%  -->
<!--   group_by(Variable,Country) %>% -->
<!--   select_at(6:16) %>% fsd -->
<!-- ``` -->

It is important to not use *dplyr*'s `summarize` together with these functions since that would totally eliminate their speed gain. These functions are fast because they are executed only once and carry out the grouped computations in C++, whereas `summarize` will apply the function to each group in the grouped tibble. - It will also work with the fast functions, but is slower than using primitive base functions since the fast functions are S3 generic -. 

***

#### Excursus: What is Happening Behind the Scenes?
To drive this point home it is perhaps good to shed some light on what is happening behind the scenes of *dplyr* and *collapse*. Fundamentally both packages follow different computing paradigms: 

*dplyr* is an efficient implementation of the Split-Apply-Combine computing paradigm. Data is split into groups, these data-chunks are then passed to a function carrying out the computation, and finally recombined to produce the aggregated data.frame. 
<!-- The efficiency of that process depends on the efficiency of the grouping, splitting, the function(s) applied and the recombining.  -->
This modus operandi is evident in the grouping mechanism of *dplyr*. When a data.frame is passed through *group_by*, a 'groups' attribute is attached: 

```{r}
GGDC10S %>% group_by(Variable,Country) %>% attr("groups")
```

This object is a data.frame giving the unique groups and in the third (last) column vectors containing the indices of the rows belonging to that group. A command like `summarize` uses this information to split the data.frame into groups which are then passed sequentially to the function used and later recombined. These steps are also done in C++ which makes *dplyr* quite efficient.  

Now *collapse* is based around one-pass grouped computations at the C++ level using its own grouped statistical functions. In other words the data is not split and recombined at all but the entire computation is performed in a single C++ loop running through that data and completing the computations for each group simultaneously. This modus operandi is also evident in *collapse* grouping objects. The method `GRP.grouped_df` takes a *dplyr* grouping object from a grouped tibble and efficiently converts it to a *collapse* grouping object: 

```{r}
GGDC10S %>% group_by(Variable,Country) %>% GRP %>% str
```

This object is a list where the first three elements give the number of groups, the group-id to which each row belongs and a vector of group-sizes. A function like `fsum` uses this information to (for each column) create a result vector of size 'N.groups' and the run through the column using the 'group.id' vector to add the i'th data point to the 'group.id[i]'th element of the result vector. When the loop is finished, the grouped computation is also finished. 

It is thus clear that *collapse* is faster than *dplyr* since it's method of computing involves less steps. 
<!-- This performance gain is realized especially as data become large, since the conversion perfomed by `GRP.grouped_df` also involves a small computational cost.  -->

***

### 1.2 More Speed using *collapse* Verbs
*collapse* fast functions do not develop their maximal performance on a grouped tibble created with `group_by` because of the additional conversion cost of the grouping object incurred by `GRP.grouped_df`. This cost is already minimized through the use of C++, but we can do even better replacing `group_by` with `collapse::fgroup_by`. `fgroup_by` works like `group_by` but does the grouping with `collapse::GRP` (up to 10x faster than `group_by`) and simply attaches a *collapse* grouping object to the grouped_df. Thus the speed gain is 2-fold: Faster grouping and no conversion cost when calling *collapse* functions.

Another improvement comes from replacing the *dplyr* verb `select` with `collapse::fselect`, and, for selection using column names, indices or functions use `collapse::get_vars` instead of `select_at` or `select_if`. Next to `get_vars`, *collapse* also introduces the predicates `num_vars`, `cat_vars`, `char_vars`, `fact_vars`, `logi_vars` and `Date_vars` to efficiently select columns by type.

```{r}
GGDC10S %>% fgroup_by(Variable,Country) %>% get_vars(6:16) %>% fmedian

microbenchmark(collapse = GGDC10S %>% fgroup_by(Variable,Country) %>% get_vars(6:16) %>% fmedian,
               hybrid = GGDC10S %>% group_by(Variable,Country) %>% select_at(6:16) %>% fmedian,
               dplyr = GGDC10S %>% group_by(Variable,Country) %>% select_at(6:16) %>% summarise_all(median, na.rm = TRUE))
```
Benchmarks on the different components of this code and with larger data are provided under 'Benchmarks'. I note that a grouped tibble created with `fgroup_by` can no longer be used for grouped computations with *dplyr* verbs like `mutate` or `summarize`. To avoid errors with these functions and `print.grouped_df`, `[.grouped_df` etc., the classes assigned after `fgroup_by` are reshuffled, so that the data.frame is treated by the *dplyr* ecosystem like a normal tibble:

```{r}
class(group_by(GGDC10S, Variable, Country))

class(fgroup_by(GGDC10S, Variable, Country))
```

I also note that `fselect` and `get_vars` are not full drop-in replacements for `select` because they do not have a grouped_df method:

```{r}
GGDC10S %>% group_by(Variable, Country) %>% select_at(6:16) %>% head(3)
GGDC10S %>% group_by(Variable, Country) %>% get_vars(6:16) %>% head(3)
```

Since by default `keep.group_vars = TRUE` in the *Fast Statistical Functions*, the end result is nevertheless the same:

```{r}
GGDC10S %>% group_by(Variable, Country) %>% select_at(6:16) %>% fmean %>% head(3)
GGDC10S %>% group_by(Variable, Country) %>% get_vars(6:16) %>% fmean %>% head(3)
```

Another useful verb introduced by *collapse* is `fgroup_vars`, which can be used to efficiently obtain the grouping columns or grouping variables from a grouped tibble:

```{r}
# fgroup_by fully supports grouped tibbles created with group_by or fgroup_by: 
GGDC10S %>% group_by(Variable, Country) %>% fgroup_vars %>% head(3)
GGDC10S %>% fgroup_by(Variable, Country) %>% fgroup_vars %>% head(3)

# The other possibilities:
GGDC10S %>% group_by(Variable, Country) %>% fgroup_vars("unique") %>% head(3)
GGDC10S %>% group_by(Variable, Country) %>% fgroup_vars("names")
GGDC10S %>% group_by(Variable, Country) %>% fgroup_vars("indices")
GGDC10S %>% group_by(Variable, Country) %>% fgroup_vars("named_indices")
GGDC10S %>% group_by(Variable, Country) %>% fgroup_vars("logical")
GGDC10S %>% group_by(Variable, Country) %>% fgroup_vars("named_logical")
```

A final *collapse* verb I want to mention here is `fsubset`, a faster alternative to `dplyr::filter` which also provides an option to flexibly subset columns after the select argument:

```{r}
# Two equivalent calls, the first is substantially faster
GGDC10S %>% fsubset(Variable == "VA" & Year > 1990, Country, Year, AGR:GOV) %>% head(3)

GGDC10S %>% filter(Variable == "VA" & Year > 1990) %>% select(Country, Year, AGR:GOV) %>% head(3)
```

### 1.3 Multi-Function Aggregations

One can also aggregate with multiple functions at the same time. For such operations it is often necessary to use curly braces `{` to prevent first argument injection so that `%>% cbind(FUN1(.), FUN2(.))` does not evaluate as `%>% cbind(., FUN1(.), FUN2(.))`:

```{r}
GGDC10S %>%
  fgroup_by(Variable,Country) %>%
  get_vars(6:16) %>% {
    cbind(fmedian(.),
          add_stub(fmean(., keep.group_vars = FALSE), "mean_"))
    } %>% head(3)
```

The function `add_stub` used above is a *collapse* function adding a prefix (default) or suffix to variables names. The *collapse* predicate `add_vars` provides a more efficient alternative to `cbind.data.frame`. The idea here is 'adding' variables to the data.frame in the first argument i.e. the attributes of the first argument are preserved, so the expression below still gives a tibble instead of a data.frame:


<!-- A slightly more elegant solution to such multi-function aggregations can be found using `get_vars`, a collapse predicate to efficiently select variables. In contrast to `select_at`, `get_vars` does not automatically add the grouping columns to the selection. -->


```{r}
GGDC10S %>%
  fgroup_by(Variable,Country) %>% {
   add_vars(ffirst(get_vars(., "Reg", regex = TRUE)),        # Regular expression matching column names
            add_stub(fmean(num_vars(.), keep.group_vars = FALSE), "mean_"), # num_vars selects all numeric variables
            add_stub(fmedian(fselect(., PU:TRA), keep.group_vars = FALSE), "median_"), 
            add_stub(fmin(fselect(., PU:CON), keep.group_vars = FALSE), "min_"))      
  }
```

Another nice feature of `add_vars` is that it can also very efficiently reorder columns i.e. bind columns in a different order than they are passed. This can be done by simply specifying the positions the added columns should have in the final data.frame, and then `add_vars` shifts the first argument columns to the right to fill in the gaps.

```{r}
GGDC10S %>%
  fsubset(Variable == "VA", Country, AGR, SUM) %>% 
  fgroup_by(Country) %>% {
   add_vars(fgroup_vars(.,"unique"),
            add_stub(fmean(., keep.group_vars = FALSE), "mean_"),
            add_stub(fsd(., keep.group_vars = FALSE), "sd_"), 
            pos = c(2,4,3,5))
  } %>% head(3)
```

A much more compact solution to multi-function and multi-type aggregation with *dplyr* is offered by the function *collapg*:

```{r}
# This aggregates numeric colums using the mean (fmean) and categorical columns with the mode (fmode)
GGDC10S %>% fgroup_by(Variable,Country) %>% collapg
```

By default it aggregates numeric columns using the `fmean` and categorical columns using `fmode`, and preserves the order of all columns. Changing these defaults is very easy:

```{r}
# This aggregates numeric colums using the median and categorical columns using the first value
GGDC10S %>% fgroup_by(Variable,Country) %>% collapg(fmedian, flast)
```

One can apply multiple functions to both numeric and/or categorical data:

```{r}
GGDC10S %>% fgroup_by(Variable,Country) %>%
  collapg(list(fmean, fmedian), list(first, fmode, flast)) %>% head(3)
```

Applying multiple functions to only numeric (or only categorical) data allows return in a long format:
```{r}
GGDC10S %>% fgroup_by(Variable,Country) %>%
  collapg(list(fmean, fmedian), cols = is.numeric, return = "long")
```

Finally, `collapg` also makes it very easy to apply aggregator functions to certain columns only:

```{r}
GGDC10S %>% fgroup_by(Variable,Country) %>%
  collapg(custom = list(fmean = 6:8, fmedian = 10:12))
```
 To understand more about `collapg`, look it up in the documentation (`?collapg`).

### 1.4 Weighted Aggregations

Weighted aggregations are currently possible with the functions `fsum, fprod, fmean, fmode, fvar` and `fsd`. The implementation is such that by default (option `keep.w = TRUE`) these functions also aggregate the weights, so that further weighted computations can be performed on the aggregated data. `fsum, fmean`, `fsd` and `fvar` compute a grouped sum of the weight column and place it next to the group-identifiers, `fmode` computes the maximum weight (corresponding to the mode), and `fprod` computes the product of the weights.

```{r}
# This computes a frequency-weighted grouped standard-deviation, taking the total EMP / VA as weight
GGDC10S %>%
  fgroup_by(Variable,Country) %>%
  fselect(AGR:SUM) %>% fsd(SUM)

# This computes a weighted grouped mode, taking the total EMP / VA as weight
GGDC10S %>%
  fgroup_by(Variable,Country) %>%
  fselect(AGR:SUM) %>% fmode(SUM)
```

The weighted variance / standard deviation is currently only implemented with frequency weights. Reliability weights may be implemented in a future update of *collapse*, if this is a strongly requested feature.

Weighted aggregations may also be performed with `collapg`.

```{r}
# This aggregates numeric colums using the weighted mean and categorical columns using the weighted mode
GGDC10S %>% group_by(Variable,Country) %>% collapg(w = SUM, wFUN = list(fsum, fmax))
```

<!-- Thus to aggregate the entire data and save the weights one would need to opt for a manual solution: -->

<!-- ```{r} -->
<!-- GGDC10S %>% -->
<!--   fgroup_by(Variable,Country) %>% { -->
<!--     add_vars(fmean(get_vars(., 6:16), SUM), -->
<!--              fmode(get_vars(., c(2:3,16)), SUM, keep.group_vars = FALSE), -->
<!--              pos = c(5, 2:3)) -->
<!--   } -->
<!-- ``` -->
<!-- <!-- ```{r} --> 
<!-- <!-- GGDC10S %>%  --> 
<!-- <!--   group_by(Variable,Country) %>% collapg(w = .$SUM) --> 
<!-- ``` -->

## 2. Fast Transformations

*collapse* also provides some fast transformations that significantly extend in scope and speed up manipulations that can be performed with `dplyr::mutate`. 
<!-- bring to *dplyr* in terms of grouped transformations. -->

### 2.1 Fast Transform and Compute Variables
The function `ftransform` can be used to manipulate columns in the same ways as `mutate`:

```{r}
GGDC10S %>% fsubset(Variable == "VA", Country, Year, AGR, SUM) %>%
  ftransform(AGR_perc = AGR / SUM * 100,  # Computing % of VA in Agriculture
             AGR_mean = fmean(AGR),       # Average Agricultural VA
             AGR = NULL, SUM = NULL) %>%  # Deleting columns AGR and SUM
             head
```

If only the computed columns need to be returned, `fcompute` provides an efficient alternative:

```{r}
GGDC10S %>% fsubset(Variable == "VA", Country, Year, AGR, SUM) %>%
  fcompute(AGR_perc = AGR / SUM * 100,
           AGR_mean = fmean(AGR)) %>% head
```

`ftransform` and `fcompute` are an order of magnitude faster than `mutate`, but they do not support grouped computations. For common grouped operations like replacing and sweeping out statistics, *collapse* however provides very efficient alternatives...

### 2.2 Replacing and Sweeping out Statistics
<!-- using Fast Statistical Functions -->

All statistical (scalar-valued) functions in the collapse package (`fsum, fprod, fmean, fmedian, fmode, fvar, fsd, fmin, fmax, ffirst, flast, fNobs, fNdistinct`) have a `TRA` argument which can be used to efficiently transforms data by either (column-wise) replacing data values with computed statistics or sweeping the statistics out of the data. Operations can be specified using either an integer or quoted operator / string. The 10 operations supported by `TRA` are:

* 1 - "replace_fill" : replace and overwrite missing values (same as `mutate`)

* 2 - "replace" : replace but preserve missing values

* 3 - "-" : subtract (center)

* 4 - "-+" : subtract group-statistics but add average of group statistics

* 5 - "/" : divide (scale)

* 6 - "%" : compute percentages (divide and multiply by 100)

* 7 - "+" : add

* 8 - "*" : multiply

* 9 - "%%" : modulus

* 10 - "-%%" : subtract modulus

<!-- For functions supporting weights (`fsum, fprod, fmean, fmode, fvar` and `fsd`) the `TRA` argument is in the third position following the data and weight vector (in the *grouped_df* method), whereas functions not supporting weights have the argument in the second position. -->

Simple transformations are again straightforward to specify:
```{r}
# This subtracts the median value from all data points i.e. centers on the median
GGDC10S %>% num_vars %>% fmedian(TRA = "-") %>% head

# This replaces all data points with the mode
GGDC10S %>% char_vars %>% fmode(TRA = "replace") %>% head
```

We can also easily specify code to grouped demean, scale or compute percentages by groups:

<!-- ^[100% being the median of all annual VA/EMP values in a given sector and country across all years, not the sectoral output share which would have to be obtained using `sweep(GGDC10S[6:16], 1, GGDC10S$SUM, "/")`] -->


```{r}
# Demeaning sectoral data by Variable and Country (within transformation)
GGDC10S %>%
  fselect(Variable,Country,AGR:SUM) %>% 
   fgroup_by(Variable,Country) %>% fmean(TRA = "-") %>% head(3)

# Scaling sectoral data by Variable and Country
GGDC10S %>%
  fselect(Variable,Country,AGR:SUM) %>% 
   fgroup_by(Variable,Country) %>% fsd(TRA = "/") %>% head(3)

# Normalizing Data by expressing them in percentages of the median value within each country and sector (i.e. the median is 100%)
GGDC10S %>%
  fselect(Variable,Country,AGR:SUM) %>%  
   fgroup_by(Variable,Country) %>% fmedian(TRA = "%") %>% head(3)

```

Weighted demeaning and scaling can be computed using:

```{r}
# Weighted demeaning (within transformation), weighted by SUM
GGDC10S %>%
  fselect(Variable,Country,AGR:SUM) %>% 
   fgroup_by(Variable,Country) %>% fmean(SUM, "-") %>% head(3)

# Weighted scaling, weighted by SUM
GGDC10S %>%
  fselect(Variable,Country,AGR:SUM) %>% 
   fgroup_by(Variable,Country) %>% fsd(SUM, "/") %>% head(3)
```

Alternatively we could also replace data points with their groupwise weighted mean or standard deviation:

```{r}
# This conducts a weighted between transformation (replacing with weighted mean)
GGDC10S %>%
  fselect(Variable,Country,AGR:SUM) %>% 
   fgroup_by(Variable,Country) %>% fmean(SUM, "replace")

# This also replaces missing values in each group
GGDC10S %>%
  fselect(Variable,Country,AGR:SUM) %>% 
   fgroup_by(Variable,Country) %>% fmean(SUM, "replace_fill")

```

<!-- It is also possible to center data points on the overall mean, which is achieved by subtracting out group means and adding the overall mean of the data again: -->
<!-- ```{r} -->
<!-- # This group-centers data on the overall mean of the data -->
<!-- GGDC10S %>% -->
<!--   group_by(Variable,Country) %>% -->
<!--     select_at(6:16) %>% fmean(TRA = "-+") -->
<!-- ``` -->
Sequential operations are also easily performed:
```{r}
# This scales and then subtracts the median
GGDC10S %>%
  fselect(Variable,Country,AGR:SUM) %>% 
   fgroup_by(Variable,Country) %>% fsd(TRA = "/") %>% fmedian(TRA = "-")
```

Of course it is also possible to combine multiple functions as in the aggregation section, or to add variables to existing data, as shown below:

```{r}
# This adds a groupwise observation count next to each column
add_vars(GGDC10S, seq(7,27,2)) <- GGDC10S %>%
    fgroup_by(Variable,Country) %>% fselect(AGR:SUM) %>%
    fNobs("replace_fill") %>% add_stub("N_")

head(GGDC10S)
rm(GGDC10S)
```


Certainly There are lots of other examples one could construct using the 10 operations and 13 functions listed above, the examples provided just outline the suggested programming basics.

<!-- ***could add add_vars example again*** -->

### 2.3 More Control using the `TRA` Function

Behind the scenes of the `TRA = ...` argument, the fast functions first compute the grouped statistics on all columns of the data, and these statistics are then directly fed into a C++ function that uses them to replace or sweep them out of data points in one of the 10 ways described above. This function can however also be called directly by the name of `TRA` (shorthand for 'transforming' data by replacing or sweeping out statistics). Fundamentally, `TRA` is a generalization of `base::sweep` for column-wise grouped operations^[Row-wise operations are not supported by TRA.]. Direct calls to `TRA` enable more control over inputs and outputs.

The two operations below are equivalent, although the first is slightly more efficient as it only requires one method dispatch and one check of the inputs:

```{r}
# This divides by the product
GGDC10S %>%
  fgroup_by(Variable,Country) %>%
    get_vars(6:16) %>% fprod(TRA = "/") 

# Same thing
GGDC10S %>%
  fgroup_by(Variable,Country) %>%
    get_vars(6:16) %>% TRA(fprod(., keep.group_vars = FALSE), "/") # [same as TRA(.,fprod(., keep.group_vars = FALSE),"/")]
```

`TRA.grouped_df` was designed such that it matches the columns of the statistics (aggregated columns) to those of the original data, and only transforms matching columns while returning the whole data.frame. Thus it is easily possible to only apply a transformation to the first two sectors:

```{r}
# This only demeans Agriculture (AGR) and Mining (MIN)
GGDC10S %>%
  fgroup_by(Variable,Country) %>%
    get_vars(6:16) %>% TRA(fmean(fselect(., AGR, MIN), keep.group_vars = FALSE), "-")
```

Another potential use of `TRA` is to do computations in two- or more steps, for example if both aggregated and transformed data are needed, or if computations are more complex and involve other manipulations in-between the aggregating and sweeping part:

```{r}
# Get grouped tibble
gGGDC <- GGDC10S %>% fgroup_by(Variable,Country)

# Get aggregated data
gsumGGDC <- gGGDC %>% fselect(AGR:SUM) %>% fsum
head(gsumGGDC)

# Get transformed (scaled) data
head(TRA(gGGDC, gsumGGDC, "/"))
```


I have already noted above that whether using the argument to fast statistical functions or `TRA` directly, these data transformations are essentially a two-step process: Statistics are first computed and then used to transform this original data. This process is already very efficient since all functions are written in C++, and programmatically separating the computation of statistics and data transformation tasks allows for unlimited combinations and drastically simplifies the code base of this package.

Nonetheless there are of course more memory efficient and faster ways to program such data transformations, which principally involve doing them column-by-column with a single C++ function. To ensure that this package lives up to the highest standards of performance for common uses, I have implemented such slightly more efficient algorithms for the very commonly applied tasks of centering and averaging data by groups (widely known as 'between'-group and 'within'-group transformations), and scaling and centering data by groups (also known as 'standardizing' data).

### 2.4 Faster Centering, Averaging and Standardizing
<!-- Between and Within Transformations and Standardization -->

The functions `fbetween` and `fwithin` are slightly more memory efficient implementations of `fmean` invoked with different `TRA` options:

```{r}
GGDC10S %>% # Same as ... %>% fmean(TRA = "replace")
  fgroup_by(Variable,Country) %>% get_vars(6:16) %>% fbetween %>% head(2)

GGDC10S %>% # Same as ... %>% fmean(TRA = "replace_fill")
  fgroup_by(Variable,Country) %>% get_vars(6:16) %>% fbetween(fill = TRUE) %>% head(2)

GGDC10S %>% # Same as ... %>% fmean(TRA = "-")
  fgroup_by(Variable,Country) %>% get_vars(6:16) %>% fwithin %>% head(2)
```

Apart from higher speed,`fwithin` has a `mean` argument to assign an arbitrary mean to centered data, the default being `mean = 0`. A very common choice for such an added mean is just the overall mean of the data, which can be added in by invoking `mean = "overall.mean"`: 

```{r}
GGDC10S %>% 
  fgroup_by(Variable,Country) %>% 
    fselect(Country, Variable, AGR:SUM) %>% fwithin(mean = "overall.mean")
```
<!-- in particular, which regards the joint use of weights and the `mean = "overall.mean"` option: `... %>% fmean(w = SUM, TRA = "-+")` will not properly group-center the data on the overall weighted mean. Instead, it will group-center data on a frequency weighted average of the weighted group-means, thus not taking into account different aggregated weights attached to those weighted group-means themselves. The reason for this shortcoming is simply that `TRA` was not designed to take a separate weight vector as input. `fwithin(w = SUM, mean = "overall.mean")` does a better job and properly centers data on the weighted overall mean after subtracting out weighted group means: -->

<!-- ```{r} -->
<!-- GGDC10S %>% # This does not center data on a properly computed weighted overall mean -->
<!--   group_by(Variable,Country) %>% select_at(6:16) %>% fmean(SUM, TRA = "-+") -->

<!-- GGDC10S %>% # This does a proper job by both subtracting weighted group-means and adding a weighted overall mean -->
<!--   group_by(Variable,Country) %>% select_at(6:16) %>% fwithin(SUM, mean = "overall.mean") -->
<!-- ``` -->

This can also be done using weights. The code below uses the `SUM` column as weights, and then for each variable and each group subtracts out the weighted mean, and then adds the overall weighted column mean back to the centered columns. The `SUM` column is just kept as it is and added in front.  

```{r}
GGDC10S %>% 
  fgroup_by(Variable,Country) %>% 
    fselect(Country, Variable, AGR:SUM) %>% fwithin(SUM, mean = "overall.mean")
```

Apart from `fbetween` and `fwithin`, the function `fscale` exists to efficiently scale and center data, to avoid sequential calls such as `... %>% fsd(TRA = "/") %>% fmean(TRA = "-")` shown in an earlier example.  

```{r}
# This efficiently scales and centers (i.e. standardizes) the data
GGDC10S %>%
  fgroup_by(Variable,Country) %>%
    fselect(Country, Variable, AGR:SUM) %>% fscale
```

`fscale` also has additional `mean` and `sd` arguments allowing the user to (group-) scale data to an arbitrary mean and standard deviation. Setting `mean = FALSE` just scales the data but preserves the means, and is thus different from `fsd(..., TRA = "/")` which just divides all values by the standard deviation:

```{r}
# Saving grouped tibble
gGGDC <- GGDC10S %>%
  fgroup_by(Variable,Country) %>%
    fselect(Country, Variable, AGR:SUM)

# Original means
head(fmean(gGGDC)) 

# Mean Preserving Scaling
head(fmean(fscale(gGGDC, mean = FALSE)))
head(fsd(fscale(gGGDC, mean = FALSE)))
```

One can also set `mean = "overall.mean"`, which group-centers columns on the overall mean as illustrated with `fwithin`. Another interesting option is setting `sd = "within.sd"`. This group-scales data such that every group has a standard deviation equal to the within-standard deviation of the data:

```{r}
# Just using VA data for this example
gGGDC <- GGDC10S %>%
  fsubset(Variable == "VA", Country, AGR:SUM) %>% 
      fgroup_by(Country)

# This calculates the within- standard deviation for all columns
fsd(num_vars(ungroup(fwithin(gGGDC))))

# This scales all groups to take on the within- standard deviation while preserving group means 
fsd(fscale(gGGDC, mean = FALSE, sd = "within.sd"))

```

A grouped scaling operation with both `mean = "overall.mean"` and `sd = "within.sd"` thus efficiently achieves a complete harmonization of all groups in the first two moments without changing the fundamental properties (in terms of level and scale) of the data. 


### 2.5 Lags / Leads, Differences and Growth Rates

<!-- It was suggested some time ago that leaving the best wine for the end is not the best strategy when giving a feast. Considering the marriage of *collapse* and *dplyr* the 3 functions for time-computations introduced in this section combine great flexibility with precision and computing power, and feature amongst the highlights of *collapse*. -->

This section introduces 3 further powerful *collapse* functions: `flag`, `fdiff` and `fgrowth`. The first function, `flag`, efficiently computes sequences of fully identified lags and leads on time-series and panel-data. The following code computes 1 fully-identified panel-lag and 1 fully identified panel-lead of each variable in the data:

<!-- In addition: None of these functions require the data to be sorted, they can carry out fast computations on completely unordered data as long as a time-variable is supplied that uniquely identifies the data. -->
```{r}
GGDC10S %>%
  fselect(-Region, -Regioncode) %>% 
    fgroup_by(Variable,Country) %>% flag(-1:1, Year)
```

If the time-variable passed does not exactly identify the data (i.e. because of gaps or repeated values in each group), all 3 functions will issue appropriate error messages. It is also possible to omit the time-variable if one is certain that the data is sorted:
```{r}
GGDC10S %>%
  fselect(Variable,Country,AGR:SUM) %>% 
    fgroup_by(Variable,Country) %>% flag
```

`fdiff` computes sequences of lagged-leaded and iterated differences as well as quasi-differences and log-differences on time-series and panel-data. The code below computes the 1 and 10 year first and second differences of each variable in the data:
```{r}
GGDC10S %>%
  fselect(-Region, -Regioncode) %>% 
    fgroup_by(Variable,Country) %>% fdiff(c(1, 10), 1:2, Year)
```
Log-differences of the form $log(x_t) - log(x_{t-s})$ are also easily computed, although one caveat of log-differencing in C++ is that `log(NA) - log(NA)` gives a `NaN` value. 

```{r}
GGDC10S %>%
  fselect(-Region, -Regioncode) %>% 
    fgroup_by(Variable,Country) %>% fdiff(c(1, 10), 1, Year, logdiff = TRUE)
```

Finally, it is also possible to compute quasi-differences and quasi-log-differences of the form $x_t - \rho x_{t-s}$ or $log(x_t) - \rho log(x_{t-s})$:

```{r}
GGDC10S %>%
  fselect(-Region, -Regioncode) %>% 
    fgroup_by(Variable,Country) %>% fdiff(t = Year, rho = 0.95)
```

The quasi-differencing feature was added to `fdiff` to facilitate the preparation of time-series and panel data for least-squares estimations suffering from serial correlation following Cochrane & Orcutt (1949). 

<!-- and `fgrowth` computes lagged-leaded and iterated growth-rates obtained via the exact computation method or through log-differencing.  -->

Finally, `fgrowth` computes growth rates in the same way. By default exact growth rates are computed in percentage terms using $(x_t-x_{t-s}) / x_{t-s} \times 100$ (the default argument is `scale = 100`). The user can also request growth rates obtained by log-differencing using $log(x_t/ x_{t-s}) \times 100$. 
```{r}
# Exact growth rates, computed as: (x - lag(x)) / lag(x) * 100
GGDC10S %>%
  fselect(-Region, -Regioncode) %>% 
    fgroup_by(Variable,Country) %>% fgrowth(c(1, 10), 1, Year)

# Log-difference growth rates, computed as: log(x / lag(x)) * 100
GGDC10S %>%
  fselect(-Region, -Regioncode) %>% 
    fgroup_by(Variable,Country) %>% fgrowth(c(1, 10), 1, Year, logdiff = TRUE)
```

`fdiff` and `fgrowth` can also perform leaded (forward) differences and growth rates, although I have never come to employ these in my personal work (i.e. `... %>% fgrowth(-c(1, 10), 1:2, Year)` would compute one and 10-year leaded first and second differences). Again it is possible to perform sequential operations:

```{r}
# This computes the 1 and 10-year growth rates, for the current period and lagged by one period
GGDC10S %>%
  fselect(-Region, -Regioncode) %>% 
    fgroup_by(Variable,Country) %>% fgrowth(c(1, 10), 1, Year) %>% flag(0:1, Year)
```

## 3. Benchmarks

This section seeks to demonstrate that the functionality introduced in the preceeding 2 sections indeed produces code that evaluates substantially faster than native *dplyr*. 

To do this properly, the different components of a typical piped call (selecting / subsetting, grouping, and performing some computation) are bechmarked separately on 2 different data sizes.

All benchmarks are run on a Windows 8.1 laptop with a 2x 2.2 GHZ Intel i5 processor, 8GB DDR3 RAM and a Samsung 850 EVO SSD hard drive.

### 3.1 Data 
Bechmarks are run on the original `GGDC10S` data used throughout this vignette and a larger dataset with approx. 1 million observations, obtained by replicating and row-binding `GGDC10S` 200 times while maintaining unique groups.

```{r, eval=NCRAN}
# This shows the groups in GGDC10S
GRP(GGDC10S, ~ Variable + Country)

# This replicates the data 200 times 
data <- replicate(200, GGDC10S, simplify = FALSE) 
# This function adds a number i to the country and variable columns of each dataset
uniquify <- function(x, i) `get_vars<-`(x, c(1,4), value = lapply(unclass(x)[c(1,4)], paste0, i))
# Making datasets unique and row-binding them
data <- unlist2d(Map(uniquify, data, as.list(1:200)), idcols = FALSE)
dim(data)

# This shows the groups in the replicated data
GRP(data, ~ Variable + Country)

gc()
```

### 3.1 Selecting, Subsetting and Grouping

```{r, eval=NCRAN, warning=FALSE, message=FALSE}
## Selecting columns
# Small
microbenchmark(dplyr = select(GGDC10S, Country, Variable, AGR:SUM),
               collapse = fselect(GGDC10S, Country, Variable, AGR:SUM))

# Large
microbenchmark(dplyr = select(data, Country, Variable, AGR:SUM),
               collapse = fselect(data, Country, Variable, AGR:SUM))

## Subsetting columns 
# Small
microbenchmark(dplyr = filter(GGDC10S, Variable == "VA"),
               collapse = fsubset(GGDC10S, Variable == "VA"))

# Large
microbenchmark(dplyr = filter(data, Variable == "VA"),
               collapse = fsubset(data, Variable == "VA"))

## Grouping 
# Small
microbenchmark(dplyr = group_by(GGDC10S, Country, Variable),
               collapse = fgroup_by(GGDC10S, Country, Variable))

# Large
microbenchmark(dplyr = group_by(data, Country, Variable),
               collapse = fgroup_by(data, Country, Variable), times = 10)

## Computing a new column 
# Small
microbenchmark(dplyr = mutate(GGDC10S, NEW = AGR+1),
               collapse = ftransform(GGDC10S, NEW = AGR+1))

# Large
microbenchmark(dplyr = mutate(data, NEW = AGR+1),
               collapse = ftransform(data, NEW = AGR+1))

## All combined with pipes 
# Small
microbenchmark(dplyr = filter(GGDC10S, Variable == "VA") %>% 
                       select(Country, AGR:SUM) %>% 
                       mutate(NEW = AGR+1) %>%
                       group_by(Country),
               collapse = fsubset(GGDC10S, Variable == "VA", Country, AGR:SUM) %>% 
                       ftransform(NEW = AGR+1) %>%
                       fgroup_by(Country))

# Large
microbenchmark(dplyr = filter(data, Variable == "VA") %>% 
                       select(Country, AGR:SUM) %>% 
                       mutate(NEW = AGR+1) %>%
                       group_by(Country),
               collapse = fsubset(data, Variable == "VA", Country, AGR:SUM) %>% 
                       ftransform(NEW = AGR+1) %>%
                       fgroup_by(Country), times = 10)

gc()
```


### 3.1 Aggregation

```{r, eval=NCRAN, warning=FALSE, message=FALSE}
## Grouping the data
cgGGDC10S <- fgroup_by(GGDC10S, Variable, Country) %>% fselect(-Region, -Regioncode)
gGGDC10S <- group_by(GGDC10S, Variable, Country) %>% fselect(-Region, -Regioncode)
cgdata <- fgroup_by(data, Variable, Country) %>% fselect(-Region, -Regioncode)
gdata <- group_by(data, Variable, Country) %>% fselect(-Region, -Regioncode)
rm(data, GGDC10S) 
gc()

## Conversion of Grouping object: This time would be required extra in all hybrid calls 
## i.e. when calling collapse functions on data grouped with dplyr::group_by
# Small
microbenchmark(GRP(gGGDC10S))

# Large
microbenchmark(GRP(gdata))


## Sum 
# Small
microbenchmark(dplyr = summarise_all(gGGDC10S, sum, na.rm = TRUE),
               collapse = fsum(cgGGDC10S))

# Large
microbenchmark(dplyr = summarise_all(gdata, sum, na.rm = TRUE),
               collapse = fsum(cgdata), times = 10)

## Mean
# Small
microbenchmark(dplyr = summarise_all(gGGDC10S, mean.default, na.rm = TRUE),
               collapse = fmean(cgGGDC10S))

# Large
microbenchmark(dplyr = summarise_all(gdata, mean.default, na.rm = TRUE),
               collapse = fmean(cgdata), times = 10)

## Median
# Small
microbenchmark(dplyr = summarise_all(gGGDC10S, median, na.rm = TRUE),
               collapse = fmedian(cgGGDC10S))

# Large
microbenchmark(dplyr = summarise_all(gdata, median, na.rm = TRUE),
               collapse = fmedian(cgdata), times = 2)

## Standard Deviation
# Small
microbenchmark(dplyr = summarise_all(gGGDC10S, sd, na.rm = TRUE),
               collapse = fsd(cgGGDC10S))

# Large
microbenchmark(dplyr = summarise_all(gdata, sd, na.rm = TRUE),
               collapse = fsd(cgdata), times = 2)

## Maximum
# Small
microbenchmark(dplyr = summarise_all(gGGDC10S, max, na.rm = TRUE),
               collapse = fmax(cgGGDC10S))

# Large
microbenchmark(dplyr = summarise_all(gdata, max, na.rm = TRUE),
               collapse = fmax(cgdata), times = 10)

## First Value
# Small
microbenchmark(dplyr = summarise_all(gGGDC10S, first),
               collapse = ffirst(cgGGDC10S, na.rm = FALSE))

# Large
microbenchmark(dplyr = summarise_all(gdata, first),
               collapse = ffirst(cgdata, na.rm = FALSE), times = 10)

## Number of Distinct Values
# Small
microbenchmark(dplyr = summarise_all(gGGDC10S, n_distinct, na.rm = TRUE),
               collapse = fNdistinct(cgGGDC10S))

# Large
microbenchmark(dplyr = summarise_all(gdata, n_distinct, na.rm = TRUE),
               collapse = fNdistinct(cgdata), times = 5)

gc()
```

<!-- The benchmarks show that at this data size efficient primitives like `base::sum` or `base::max` can still deliver very decent performance with `summarize`. Less optimized base functions like `mean`, `median` and `sd` however take multiple seconds to compute, and here `collapse` fast functions really prove to be very useful complements to the *dplyr* system. -->

<!-- Weighted statistics are also performed extremely fast by *collapse* functions. I would not know how to compute weighted statistics by groups in *dplyr*, as it would require the weighting variable to be split as well, which seems impossible in native *dplyr*. -->

<!-- A further highlight of *collapse* is the extremely fast statistical mode function, which can also compute a weighted mode. Fast categorical aggregation has been an issue in R, and defining a mode function from base R and applying it to 17000 groups will probably let it run at least a minute. `fmode` reduces this time to half a second. -->

<!-- Thus in terms of data aggregation *collapse* fast functions are able to speed up *dplyr* to a level that makes it attractive again to R users working on medium-sized or larger data, and everyone programming with *dplyr*. I however strongly recommend *collapse* itself for easy and speedy programming as it does not rely on non-standard evaluation and has less R-overhead than *dplyr*. -->

<!-- In all of this the grouping system of *dplyr* remains the central bottleneck. For example grouping 10 million observations in 1 million groups takes around 10 second with `group_by`, whereas `GRP` takes around 1.5 seconds, and this difference grows exponentially as data get larger. Rewriting `group_by` using `GRP` / *data.table*'s `forderv` and then writing a simple C++ conversion program for the grouping object could be a quick remedy for this issue, but that is at the discretion of Hadley Wickham and coauthors. -->
<!-- (If you need that speed program with *collapse* or use *data.table* with GeForce optimized functions). -->

Below I add in some benchmarks for weighted aggregations and aggregations using the statistical mode, which cannot easily or efficiently be performed with *dplyr*. 

```{r, eval=NCRAN, warning=FALSE, message=FALSE}
## Weighted Mean
# Small
microbenchmark(fmean(cgGGDC10S, SUM)) 

# Large 
microbenchmark(fmean(cgdata, SUM), times = 10) 

## Weighted Standard-Deviation
# Small
microbenchmark(fsd(cgGGDC10S, SUM)) 

# Large 
microbenchmark(fsd(cgdata, SUM), times = 10) 

## Statistical Mode
# Small
microbenchmark(fmode(cgGGDC10S)) 

# Large 
microbenchmark(fmode(cgdata), times = 10) 

## Weighted Statistical Mode
# Small
microbenchmark(fmode(cgGGDC10S, SUM)) 

# Large 
microbenchmark(fmode(cgdata, SUM), times = 10) 

gc()
```

### 3.2 Transformation

```{r, eval=NCRAN, warning=FALSE, message=FALSE}

## Replacing with group sum
# Small
microbenchmark(dplyr = mutate_all(gGGDC10S, sum, na.rm = TRUE),
               collapse = fsum(cgGGDC10S, TRA = "replace_fill"))

# Large
microbenchmark(dplyr = mutate_all(gdata, sum, na.rm = TRUE),
               collapse = fsum(cgdata, TRA = "replace_fill"), times = 10)

## Dividing by group sum
# Small
microbenchmark(dplyr = mutate_all(gGGDC10S, function(x) x/sum(x, na.rm = TRUE)),
               collapse = fsum(cgGGDC10S, TRA = "/"))

# Large
microbenchmark(dplyr = mutate_all(gdata, function(x) x/sum(x, na.rm = TRUE)),
               collapse = fsum(cgdata, TRA = "/"), times = 10)

## Centering
# Small
microbenchmark(dplyr = mutate_all(gGGDC10S, function(x) x-mean.default(x, na.rm = TRUE)),
               collapse = fwithin(cgGGDC10S))

# Large
microbenchmark(dplyr = mutate_all(gdata, function(x) x-mean.default(x, na.rm = TRUE)),
               collapse = fwithin(cgdata), times = 10)

## Centering and Scaling (Standardizing)
# Small
microbenchmark(dplyr = mutate_all(gGGDC10S, function(x) (x-mean.default(x, na.rm = TRUE))/sd(x, na.rm = TRUE)),
               collapse = fscale(cgGGDC10S))

# Large
microbenchmark(dplyr = mutate_all(gdata, function(x) (x-mean.default(x, na.rm = TRUE))/sd(x, na.rm = TRUE)),
               collapse = fscale(cgdata), times = 2)

## Lag
# Small
microbenchmark(dplyr_unordered = mutate_all(gGGDC10S, dplyr::lag),
               collapse_unordered = flag(cgGGDC10S),
               dplyr_ordered = mutate_all(gGGDC10S, dplyr::lag, order_by = "Year"),
               collapse_ordered = flag(cgGGDC10S, t = Year))

# Large
microbenchmark(dplyr_unordered = mutate_all(gdata, dplyr::lag),
               collapse_unordered = flag(cgdata),
               dplyr_ordered = mutate_all(gdata, dplyr::lag, order_by = "Year"),
               collapse_ordered = flag(cgdata, t = Year), times = 2)

## First-Difference (unordered)
# Small
microbenchmark(dplyr_unordered = mutate_all(gGGDC10S, function(x) x - dplyr::lag(x)),
               collapse_unordered = fdiff(cgGGDC10S))

# Large
microbenchmark(dplyr_unordered = mutate_all(gdata, function(x) x - dplyr::lag(x)),
               collapse_unordered = fdiff(cgdata), times = 2)

gc()
```

Again below I add in some benchmarks for transformations not easily of efficiently performed with *dplyr*, such as centering on the overall mean, mean-preserving scaling, weighted scaling and centering, sequences of lags / leads, (iterated) panel-differences and growth rates. 

```{r, eval=NCRAN, warning=FALSE, message=FALSE}
# Centering on overall mean
system.time(fwithin(cgdata, mean = "overall.mean"))

# Weighted Centering
system.time(fwithin(cgdata, SUM))
system.time(fwithin(cgdata, SUM, mean = "overall.mean"))

# Weighted Scaling and Standardizing
system.time(fsd(cgdata, SUM, TRA = "/"))
system.time(fscale(cgdata, SUM))

# Sequence of lags and leads
system.time(flag(cgdata, -1:1))

# Iterated difference
system.time(fdiff(cgdata, 1, 2))

# Growth Rate
system.time(fgrowth(cgdata,1))
```

<!-- Again the benchmarks show stunning performance gains using *collapse* functions. -->

```{r, echo=FALSE}
options(oldopts)
```


## References

Timmer, M. P., de Vries, G. J., & de Vries, K. (2015). "Patterns of Structural Change in Developing Countries." . In J. Weiss, & M. Tribe (Eds.), *Routledge Handbook of Industry and Development.* (pp. 65-83). Routledge.

Cochrane, D. & Orcutt, G. H. (1949). "Application of Least Squares Regression to Relationships Containing Auto-Correlated Error Terms". *Journal of the American Statistical Association.* 44 (245): 32–61. 

Prais, S. J. & Winsten, C. B. (1954). "Trend Estimators and Serial Correlation". *Cowles Commission Discussion Paper No. 383.* Chicago.

